---
title: "Visualisation with CRM in `trialr`"
author: "Kristian Brock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In the general [introductory CRM vignette](https://brockk.github.io/trialr/articles/CRM.html), we introduced the different flavours of the Continual Reassessment Method (CRM) implmented in `trialr`.
In this vignette, we demonstrate some visualisation methods that will be useful, applicable in all variants.

For illustration, let us assume that we have treated 6 patients at 3 dose-levels:

| Patient | Cohort | Dose-level | DLT |
|:-------:|:------:|:----------:|:---:|
|    1    |   1    |      2     |  0  |
|    2    |   1    |      2     |  0  |
|    3    |   2    |      3     |  0  |
|    4    |   2    |      3     |  0  |
|    5    |   3    |      4     |  1  |
|    6    |   3    |      4     |  1  |

```{r}
doses_given <- c(2, 2, 3, 3, 4, 4)
dlt <- c(0, 0, 0, 0, 1, 1)
```

and that we are using the empiric CRM with the following prior to target the dose with Pr(DLT) closest to 25%:

```{r}
skeleton <- c(0.05, 0.15, 0.25, 0.4, 0.6)
target <- 0.25
```

To access the Stan implementations of the CRM, we must load `trialr`:
```{r, message=FALSE, warning=FALSE}
library(trialr)
```

and to fit the data to the model, we run:
```{r, results = "hide", warning=FALSE, message=FALSE}
dat <- list(
  a0 = 3,
  beta_sd = sqrt(1.34),
  num_doses = length(skeleton),
  skeleton = skeleton,
  num_patients = length(doses_given),
  tox = dlt,
  doses = doses_given
)
samp <- rstan::sampling(stanmodels$CrmEmpiricNormalPrior,
                        data = dat, seed = 123)
```

Recall that we set the random number generator seed in demonstrations for reproducibility, but you would probably not do this in a real life.

We learned [before](https://brockk.github.io/trialr/articles/CRM.html) that we can easily calculate the posterior probability of toxicity at each dose:

```{r}
prob_tox_samp <- as.data.frame(samp, 'prob_tox')
colMeans(prob_tox_samp)
```

and infer that, in this instance, dose-level 2 is closest to our target toxicity rate of 25%.
However, we will probably like to convey this information to the research world in a visually-appealing manner.
This is simple with access to the posterior samples.

For ease of plotting with `ggplot2`, we recompose the posterior samples to a tall format using `tidyr` and `dplyr`:

```{r, warning=FALSE, message=FALSE}
library(tidyr)
library(dplyr)

prob_tox_samp_tall <- prob_tox_samp %>%
  gather(Label, ProbTox) %>%
  mutate(
    DoseLevel = rep(1:ncol(prob_tox_samp), each = nrow(prob_tox_samp)),
    Draw = rep(1:nrow(prob_tox_samp), times = ncol(prob_tox_samp))
  )
```

for instance:

```{r}
prob_tox_samp_tall %>% head(10)
```

Boxplots would be a traditional way of visualising the distributions of the probability of toxicity at each dose:

```{r, fig.width=7, fig.height=7}
library(ggplot2)

(p1 <- ggplot(prob_tox_samp_tall, aes(x = DoseLevel, y = ProbTox, group = DoseLevel)) +
  geom_boxplot() + ylim(0, 1) + labs(title = 'boxplot of Pr(DLT) under CRM'))
```

However, boxplots give only limited information on the distributions.
For instance, it might be tempting to assume that the probability of toxicity is normally distributed at each dose-level.
The boxplots suggest some wide tails.
This inference is much more clear, however, using a violin-plot:

```{r, fig.width=7, fig.height=7}
(p2 <- ggplot(prob_tox_samp_tall, aes(x = DoseLevel, y = ProbTox, group = DoseLevel)) +
    geom_violin(fill = 'orange') + ylim(0, 1) + labs(title = 'violin plot of Pr(DLT) under CRM'))
```

If you are a child of the nineties (and you have installed the [`ggridges`](https://cran.r-project.org/web/packages/ggridges/) package), you may however prefer to show this information using a ridge plot, aka a [joyplot](https://en.wikipedia.org/wiki/Unknown_Pleasures)

```{r, fig.width=7, fig.height=7}
library(ggridges)
ggplot(prob_tox_samp_tall %>% mutate(DoseLevel = factor(DoseLevel)),
       aes(x = ProbTox, y = DoseLevel, fill = DoseLevel)) +
  geom_density_ridges() + theme(legend.position = 'none') +
  labs(title = 'joyplot of Pr(DLT) under CRM')
```

We see that the posterior probabilities of Pr(DLT) are far from normal, under this model.

We will naturally want to visualise quantities beyond just the probability of toxicity.
We learned in the [introductory CRM vignette](https://brockk.github.io/trialr/articles/CRM.html) that with the full Bayesian CRM provided by `trialr` and `rstan`, we can calculate the probability that each dose is the maximum tolerable dose (MTD).
The natural next step is to visualise that trade-off:

```{r, fig.width=7, fig.height=7, message=FALSE, warning=FALSE}
library(magrittr)

apply(prob_tox_samp, 1, function(x) which.min(abs(x - target))) %>% 
  data.frame(Dose = .) %>% 
  ggplot(aes(x = Dose)) + geom_bar() + 
  labs(title = 'Posterior probability that each dose is the MTD',
       y = 'Prob(MTD)')
```

In this interim stage, each of the first four doses could plausibly be the MTD, but the top dose looks unlikely.
This information was not readily available from the above plots of the probabilities of toxicity.

We might also like to visualise the probability that the toxicity rate at each dose exceeds our target toxicity rate.

```{r, fig.width=7, fig.height=7}
colMeans(prob_tox_samp > target) %>% 
  data.frame(ProbToxic = ., DoseLevel = 1:length(skeleton)) %>%
  ggplot(aes(x = DoseLevel, y = ProbToxic)) + geom_col() + 
  labs(title = 'Posterior probability that each dose is too toxic',
       y = 'Prob(Tox > target)')
```




# References
