---
title: "EffTox in trialr"
author: "Kristian Brock"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
bibliography: C:/Users/krist/Dropbox/Projects/library.bib
vignette: >
  %\VignetteIndexEntry{EffTox in trialr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Recap of the EffTox model
@Thall2004 introduced the _EffTox_ dose-finding design for scenarios where both efficacy and toxicity events should guide dose selections.
Full details are given in @Thall2004, @Thall2006 and @Thall2014 but a brief recap summary follows.

For doses $\boldsymbol{y} = (y_1, ..., y_n)$, the authors define codified doses $(x_1, ..., x_n)$ using the transform 

$x_i = \log{y_i} - \sum_{j=1}^n \frac{\log{y_j}}{n}$

The codified doses are used as the sole explanatory variables in logit models for the marginal probabilities of toxicity and efficacy:

$\pi_T = \alpha + \beta x$

$\pi_E = \gamma + \zeta x + \eta x^2$

and the joint probability is modelled using the function

$\pi_{a,b} = (\pi_E)^a (1-\pi_E)^{1-a} (\pi_T)^b (1-\pi_T)^{1-b} + (-1)^{a+b} (\pi_E) (1-\pi_E) (\pi_T) (1-\pi_T) \frac{e^\psi-1}{e^\psi+1}$.

At each dose update decision, the dose $x$ is acceptable if

$\text{Pr}\left\{ \pi_T(x, \boldsymbol{\theta}) < \overline{\pi}_T | \mathcal{D} \right\} > p_T$   and   $\text{Pr}\left\{ \pi_E(x, \boldsymbol{\theta}) > \underline{\pi}_E | \mathcal{D} \right\} > p_E$

and is no more than than one position below the lowest dose-level given and no more than one position above the highest dose-level given.
The net effect of these last two criteria is that untried doses may not be skipped in escalation of de-escalation.
$\underline{\pi}_E, p_E, \overline{\pi}_T, p_T$ are provided by the user as the trial scenario dictates.

The utility of dose $x$ is 

$TODO$

The dose-level from the acceptable set with maximal utility is selected to be given to the next patient or cohort.
If there are no acceptable doses, the trial stops and no dose is recommended.

Normal priors are specified for the elements of the parameter vector $\boldsymbol{\theta} = (\alpha, \beta, \gamma, \zeta, \eta, \psi)$.

There are several published examples of EffTOx trials, including explanations and tips on parameter choices  [@Thall2004, @Thall2006, @Thall2014, TODO].

The MD Anderson Cancer Center publishes software [@EffTox] to perform calculations and simulations for EffTox trials.
However, tHe software is available for Windows in compiled-form only.
Thus, you will not be able to run the software on Mac or Linux.
Furthermore, you will not be able to alter the behaviour of the model.
Brock TODO describes a real trial scenario where some alteration of the default model behaviour would have been preferable.
It was this that prompted the author to write the open-source implementation provided in 'trialr'.

TODO

## EffTox in `trialr`
We will work with the scenario given in @Thall2014.
This is the parameterisation loaded by default in the MD Anderson EffTox app.

```{r}
library(trialr)
```
They investigate five doses 1, 2, 4, 6.6 and 10 mcL/kg, seeking the best dose with $\text{Pr}\left\{ \pi_T(x, \boldsymbol{\theta}) < 0.3 | \mathcal{D} \right\} > 0.1$ and $\text{Pr}\left\{ \pi_E(x, \boldsymbol{\theta}) > 0.5 | \mathcal{D} \right\} > 0.1$ 

thus $p_T = p_E = 0.1, \overline{\pi}_T = 0.3$ and $\underline{\pi}_E = 0.5$.

Their hinge points on the neutral utility curve are (0.5, 0), (1, 0.65) and (0.7, 0.25), thus we have

```{r}
p <- efftox_solve_p(eff0 = 0.5, tox1 = 0.65, eff_star = 0.7, tox_star = 0.25)
p
```
We see that the utility curves in this particular example have curvature parameter $p \approx 0.977$.

Let's say that we have treated six patients in two cohorts of three at dose-levels 1 and 2 respectively. 
ALso, let's say we observe no toxicities and one efficacy event in cohort 1, and one toxicity and two efficacies in cohort 2. 

```{r}
doses = c(1, 1, 1, 2, 2, 2)
tox   = c(0, 0, 0, 0, 1, 0)
eff   = c(0, 1, 0, 0, 1, 1)
```

The entire data list for this example, including hyperprior parameters, dose admissibility parameters and patient outcomes is

```{r}
dat <- list(
  num_doses = 5,
  real_doses = c(1, 2, 4, 6.6, 10),
  efficacy_hurdle = 0.5,
  toxicity_hurdle = 0.3,
  p = p,
  eff0 = 0.5,
  tox1 = 0.65,

  alpha_mean = -7.9593, alpha_sd = 3.5487,
  beta_mean = 1.5482, beta_sd = 3.5018,
  gamma_mean = 0.7367, gamma_sd = 2.5423,
  zeta_mean = 3.4181, zeta_sd = 2.4406,
  eta_mean = 0, eta_sd = 0.2,
  psi_mean = 0, psi_sd = 1,

  num_patients = 6,
  doses = c(1, 1, 1, 2, 2, 2),
  tox   = c(0, 0, 0, 0, 1, 0),
  eff   = c(0, 1, 0, 0, 1, 1)
)
```

We invoke the prior-to-posterior analysis by calling the `rstan::sampling` function with the pre-compiled EffTox stanmodel^[The advice by the RStan team is that authors of packages that use RStan should provide static model files that are compiled when the package is installed.] and our data list.
We set the seed for reproducibility.

```{r, results = "hide"}
samp <- rstan::sampling(stanmodels$EffTox, data = dat, seed = 123)
```

The call to `rstan:sampling` prints logging information to the out stream.
That has been suppressed here for brevity.

The post-processing call to `efftox_process` extracts the most pertinent information from the sampling
object.
The parameters `p_e` and `p_t` are required here to calculate posterior dose acceptability.
They were not required in the prior-to-posterior analysis, so they were not provided in `dat`.
```{r}
x <- efftox_process(dat, samp, p_e = 0.1, p_t = 0.1)
x$recommended_dose
```

We see that dose-level 3 is recommended for the next cohort.

```{r}
knitr::kable(efftox_analysis_to_df(x), digits = 3)
```

This confirms that dose-level 3 has maximal utility.

You can produce posterior density plots

```{r, fig.width = 6, fig.cap = "Posterior densities of utility."}
u <- rstan::extract(samp, 'utility')[[1]]
df <- data.frame(Utility = as.vector(u), 
                 Dose = factor(rep(1:ncol(u), each = nrow(u))))
library(ggplot2)
ggplot(df, aes(x = Utility, group = Dose, col = Dose)) + geom_density()
```



## Quoting

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))


# References
