---
title: "The `trialr` package"
author: "Kristian Brock"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`trialr` is a collection of Bayesian clinical trial designs implemented in Stan and R.

Many notable Bayesian designs for clinical trials have been published.
However, one of the factors that has constrained their adoption is availability of software.
We present here some of the most popular, implemented and demonstrated in a consistent style, leveraging the powerful Stan environment.

It should be stressed that Bayesian trialists are not generally without code.
Often authors make available code with their design publication.
There are also some fantastic packages that aid the use of certain designs.
However, challenges to use still persist.
The disparate methods are naturally presented in a style that appeals to the particular author.
Features implemented in one package for one design may be missing in another.
Sometimes the technology chosen may only be available on one particular operating system, or the chosen technology may have fallen into disuse.

`trialr` seeks to address these problems.
Models are specified in [Stan](http://mc-stan.org/), a state-of-the-art environment for Bayesian analysis.
It uses Hamiltonian Monte Carlo to take samples from the posterior distributions.
This method is more efficient than Gibbs sampling, for instance, and reliable inference can be performed on a few thousand posterior samples.
Both R and Stan are available on Mac, Linux, and Windows, and all the examples presented here should work on each operating system.
Furthermore, Stan offers a very simple method to split the sampling across _n_ cores, taking full advantage of the modern multicore processor in your computer (probably).

The designs implemented in `trialr` are introduced briefly below, and developed more fully in vignettes.
We focus on real-life usage, including:

- fitting models to observed data using your prior;
- processing posterior samples to produce useful inferences;
- and visualising inferences using modern `ggplot` graphics.

# Examples
In all examples, we will need to load `trialr`

```{r, message=FALSE, warning=FALSE}
library(trialr)
```

## CRM
The Continual Reassessment Method was first published by O'Quigley, Pepe and Fisher (1990).
It assumes a smooth mathematical form for the dose-toxicity curve to conduct a dose-finding trial seeking a maximum tolerable dose.
There are many variations to suit different clinical scenarios and the design has enjoyed _relatively_ common use (although nowhere near as common as the ubiquitous and inferior 3+3 design).

We illustrate the 


## EffTox
Thall & Cook's EffTox is a dose-finding design that uses binary efficacy and toxicity outcomes to select a dose with a high utility score. 
We present it briefly here there is a much more thorough examination in the [EffTox vignette](https://brockk.github.io/trialr/articles/EffTox.html).

We include a convenience function to get parameters for the EffTox demo scenario of a study with five doses:

```{r}
dat <- efftox_parameters_demo()
```

but generally you would specify parameters in a list. See the [EffTox vignette](https://brockk.github.io/trialr/articles/EffTox.html) for full details.

Let us add outcomes for 3 patients: 

- patient 1, treated at dose 1, has neither efficacy nor toxicity; 
- patient 2, treated at dose 2, has just efficacy; 
- and patients 3, treated at dose 3,  has both:

```{r}
dat$num_patients <- 3
dat$eff <- c(0, 1, 1)
dat$tox <- c(0, 0, 1)
dat$doses <- c(1, 2, 3)
```

To get posterior samples, we invoke Stan the model on our `dat` object. 
The EffTox model is include in `trialr` as `stanmodels$EffTox`.
The `stanmodels` object is compiled when you install `trialr` and made available when you load the package.

```{r, message=FALSE, warning=FALSE}
set.seed(123)
samp <- rstan::sampling(stanmodels$EffTox, data = dat, refresh = FALSE)
```

The `efftox_process` function is provided to perform the calculations on the posterior samples necessary to follow the design published by Thall & Cook:

```{r}
decision <- efftox_process(dat, samp)
```

For instance, after evaluation of our three patients, the dose advocated for the next group is dose-level 3: 
```{r}
decision$recommended_dose
```

This is not surprising because dose 3 has the highest utility score:

```{r}
decision$utility
```

Sometimes, doses other than the maximal-utility dose will be recommended because of the dose-admissibility rules. 
See the papers for details. 

Functions are provided to create useful plots. 
For instance, it is illuminating to plot the posterior means of the probabilities of efficacy and toxicity at each of the doses on the trade-off contours.
The five doses are shown in red. Doses closer to the lower-right corner have higher utility. 

```{r, fig.width=7, fig.height=7}
efftox_contour_plot(dat, prob_eff = decision$prob_eff, prob_tox = decision$prob_tox)
title('EffTox utility contours')
```

This example continues in the [EffTox vignette](https://brockk.github.io/trialr/articles/EffTox.html).

There are many publications related to EffTox but the two most important are _Dose-Finding Based on Efficacy-Toxicity Trade-Offs_ (Thall & Cook, 2004) and _Effective sample size for computing prior hyperparameters in Bayesian phase I-II dose-finding_ (Thall, Herrick, Nguyen, Venier & Norris, 2014).



## Hierachical analysis of response in related cohorts
TODO

## BEBOP in PePS2
TODO
